Sampling Strategies in Decisions from Experience

### 2022-01-12 manuscript.Rmd
Preliminary Introduction

For one, there is the well-known statistical *induction problem* that the sample size determines the margin of error within which any true property of a choice alternative can be inferred, given an otherwise perfect inference process. 
Yet, the underweighting pattern that is caused by the adoption of a certain selection criterion is a priori predicted to become more pronounced with increasing sample size. 
This prediction is in line with the principle already stated above, i.e., that choice patterns should become more systematic as property descriptions become more accurate.
Nevertheless, the prediction is different from the explanation that underweighting is caused by the reliance on small samples and the skewness of the binomial distribution [e.g., @foxDecisionsExperienceSampling2006; @hertwigDecisionsExperienceEffect2004; @hauDescriptionexperienceGapRisky2008; @ungemachAreProbabilitiesOverweighted2009; @rakowBiasedSamplesNot2008], highlighting that outside the *statistical* models used to describe the choice data, *causal* decision making models of quite different nature may be at play, and that a rather important difference between the latter resides in the selection criteria they include.
I will elaborate on this matter as I proceed, however, to be rather clear from the outset, what the mind does in neither case is performing an underweighting computation in its literal sense, but it processes the inputs in a way that the outputs approximate the solutions of what can be considered probability weighting on an abstract computational level [cf. @griffithsRationalUseCognitive2015].

The theoretical argument is substantiated by a simulation study. 
For this purpose, the sampling and decision making processes are expressed mathematically and implemented in a computational model.
Choice data from DfE is simulated and modeled using cumulative prospect theory (CPT), a data model commonly used for describing decisions under risk and uncertainty [@tverskyAdvancesProspectTheory1992]. 
All simulations and analyses are reproducible---materials and instructions can be found on the GitHub repository.

@@

Decision making is an information-processing problem whose solution requires a selection among choice alternatives. 
I.e., with regards to a given selection criteria, all chosen alternatives are equal to each other but different from the non-chosen alternatives.
Now, consider the power set $\mathcal{P}(I)$, which is the set of all possible subsets of $I$ (see example in Figure 2).
Each of the subsets $I' \in \mathcal{P}(I)$ corresponds to a possible distinction between the choice alternatives in $I$ that can be made after assessing for each alternative whether it fulfills the selection criteria or not [cf. @jostMathematicalConcepts2015, p. 16]. 
The concept of the power set thereby illustrates in a plain way how different selection criteria can alter the consequential choice, provided the properties are not uniformly distributed among the alternatives and the mind is sensitive to whatever relations the alternatives have with each other. 
\

>>>>

For one, there is the well-known statistical *induction problem* that the sample size determines the margin of error within which any true property of a choice alternative can be inferred, given an otherwise perfect inference process. 
Yet, the underweighting pattern that is caused by the adoption of a certain selection criterion is a priori predicted to become more pronounced with increasing sample size. 
This prediction is in line with the principle already stated above, i.e., that choice patterns should become more systematic as property descriptions become more accurate.
Nevertheless, the prediction is different from the explanation that underweighting is caused by the reliance on small samples and the skewness of the binomial distribution [e.g., @foxDecisionsExperienceSampling2006; @hertwigDecisionsExperienceEffect2004; @hauDescriptionexperienceGapRisky2008; @ungemachAreProbabilitiesOverweighted2009; @rakowBiasedSamplesNot2008], highlighting that outside the *statistical* models used to describe the choice data, *causal* decision making models of quite different nature may be at play [cf. @mcelreathStatisticalRethinkingBayesian2020], and that a rather important difference between the latter resides in the selection criteria they include.
I will elaborate on this matter as I proceed, however, to be clear from the outset, what the mind does in neither case is performing an underweighting computation in its literal sense, but it processes the inputs in a way that the outputs approximate the solutions of what can be considered probability weighting on an abstract computational level [cf. @griffithsRationalUseCognitive2015].

For another, this article approaches the problem of organizing the sampling data in such a way that it allows chains of inferences connecting the input set $I$ with the output set $I' \subset I$ that are algorithmically short and simple enough to be carried out by minds whose cognitive capacities are bounded [see @simonBehavioralModelRational1955; @simonInvariantsHumanBehavior1990]. 
This may be considered a *representation problem* since the inference chains are constructed around explicit descriptions of the properties of choice alternatives and representations are the "formal schemes" of symbols and rules with which these descriptions are derived [@marrRepresentationRecognitionSpatial1978, p. 270]. 
Specifically, representations differ in how they organize the information used to infer the descriptions of certain properties and these organizations determine the algorithms that can be built into the inference chain [@marrVisionComputationalInvestigation1982].
In this article, *sampling strategies* in DfE are considered as representation systems used to organize the samples in a way that subserves the property inferences required by the application of particular selection criteria.
Though most of my expositions on the role of sampling strategies remain speculative, the core idea is not new to decision research [see, e.g., @hillsInformationSearchDecisions2010; @wulffHowShortLongrun2015].
By restating this idea and adopting some terms and concepts that have well-defined meaning in math and cognitive science, I hope to integrate it into the rich study of the computationally limited mind as an adaptive cognitive system whose processes interact with the information structures in the environment or memory [e.g., @simonRationalChoiceStructure1956; @hertwigTamingUncertainty2019].

The theoretical arguments sketched in this admittingly dense introduction are substantiated by a simulation study. 
For this purpose, the sampling and decision making processes are expressed mathematically and implemented in a computational model to simulate choice data from DfE.
The simulated data is modeled in cumulative prospect theory (CPT), a data model commonly used for describing decisions under risk and uncertainty [@goldsteinExpressionTheoryPreference1987; see also @tverskyAdvancesProspectTheory1992]. 
All simulations and analyses are reproducible---materials and instructions can be found on the GitHub repository.

@@

We already stated that decision making is an information-processing problem whose solution requires a selection among choice alternatives.  all chosen alternatives are equal to each other but different from the non-chosen alternatives.

Now, consider the power set $\mathcal{P}(I)$, which is the set of all possible subsets of $I$ (see example in Figure 2).
Each of the subsets $I' \in \mathcal{P}(I)$ corresponds to a possible distinction between the choice alternatives in $I$ that can be made after assessing for each alternative whether it fulfills the selection criteria or not [cf. @jostMathematicalConcepts2015, p. 16]. 
The concept of the power set thereby illustrates in a plain way how different selection criteria can alter the consequential choice, provided the properties are not uniformly distributed among the alternatives and the mind is sensitive to whatever relations the alternatives have with each other.
\
